Transcriptomic responses to diet quality and viral infection in Apis mellifera

### 2018-10-26 Manuscript/Text/manuscript.tex
Added diet transcriptomics in model organisms paragraph

As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees. Honey bee transcriptomic studies have found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.

On the other hand, gene expression studies in insect models such as Drosophila and mosquito have shown that RNA silencing is a major antiviral strategy, but other pathways play substantial roles in many virus-host systems like the Toll pathway, IMD pathway, JAK/STAT pathway, Toll-7-autophagy pathway and transcriptional pausing \cite{cherryReview, swevers}. It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}. In fact, every virus-host interaction has its own particularities derived of the diverse ways of replication and infection cycle evolved by different viruses. In this sense, an intricate set of pro- and anti-virus host factors are involved in each case \cite{cherryReview, swevers}. For example, dicistroviruses Drosophila C virus (DCV) and cricket paralysis virus (CrPV) infection requires ribosomal proteins RpS6, RpL19 and RACK1, but such proteins are dispensable during infection by the rhabdovirus vesicular stomatitis virus (VSV) in Drosophila. In contrast, the autophagy pathway has proven to be an effective antiviral strategy against VSV, but has no role limiting DCV or other positive-sense RNA viruses \cite{swevers}. For an extensive up-to-date review of known antiviral and proviral factors for specific viruses in Drosophila see \cite{swevers}. Model insect studies have shed light on how honey bees respond to viruses because the function of many genes relies on the characterization of orthologue genes. Nonetheless, recent transcriptomic studies in honey bees have improved our understanding of compatible and incompatible virus-bee interactions. Key factors or hallmarks of the antiviral conserved defense pathways mentioned above have been found in some virus-bee systems (e.g. RNA silencing, Toll, IMD, JAK/STAT pathway, autophagy, and endocytosis) \cite{McMenamin}. Added to these, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken}. This path has been studied recently in more depth using transcriptomics, which found known immune factors but also new ones for which indispensability in antiviral response was determined by silencing \cite{Brutscher}. Similar to Drosophila examples, particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed. This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs. deformed wing virus (DWV) \cite{Ryabov}. Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}. In the case of dicistroviruses, few works have studied the impact of IAPV infection at transcriptional level. Chen et al. 2014 analyzed responses to IAPV infection in larvae and workers using microarrays \cite{Chen}. Many of the DEGs found were involved in immune response and energy-related metabolism, particularly in adults but not in brood. The authors propose this observed difference could be connected to latent infections in larvae (where host immunity is not perturbed) versus acute infections in adulthood (induced by stressors faced during development) \cite{Chen}. IAPV acute infection also alters the DNA methylation pattern of numerous genes that do not overlap the genes that are up- or down-regulated at the transcriptional level \cite{galbraith}. These works reiterate the conclusion that viruses trigger particular antiviral mechanisms by different means and depending on several factors. The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy to mention that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.

@@

There were several limitations in this study that could be improved upon in future studies. For instance, our comparison between the Galbraith data (single-drone colonies) and our data (polyandrous colonies) was limited by numerous extraneous variables between these studies. In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on one-day old worker honey bees that were fed sugar and artificial pollen diet, whereas our study focused on adult worker honey bees that were fed bee-collected monofloral diets. Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and only considered symptomatic honey bees for their infected treatment group, whereas we used whole bodies and considered both asymptomatic and symptomatic honey bees for our infected treatment group. Further differences between the studies can be found in their corresponding published methods sections \cite{adamInt, galbraith}. Our comparative visualization assessment between these two datasets was also somewhat limited because the virus effect in the Galbraith study used three replicates for each level, whereas the virus effect in our study used twelve replicates for each level that were actually further subdivided into six replicates for each diet level. Hence the apparent reduction in noise observed in the Galbraith data compared to our data in the PCA plots, parallel coordinate plots, scatterplot matrices, and litre plots may be an inadvertent product of the smaller number of replicates used and the lack of a secondary treatment group rather than solely the reduction in genetic variability through the single-drone colony design itself. With this in mind, while our current efforts may be a starting point, future studies can shed more light on signal:to:noise and differential expression differences between polyandrous colony designs and single-drone colony designs by controlling for extraneous factors more strictly than what we were able to do in the current line of work.

>>>>

There are only a small number of published experiments examining gene expression patterns related to diet effects \cite{alaux2} and virus infection effects \cite{galbraith} in honey bees, but there have also been several such studies in model organisms. For example, diet effects on transcriptomics have been studied in insect models such as Drosophila. For instance, one recent transcriptomic study in Drosophila melanogaster reported an overexpression of genes related to immunity, metabolism, and hemocyanin in a high-fat diet and overexpression of genes related to cell cycle activity, DNA binding and transcription, and CHK kinase-like protein activity in a high-sugar diet \cite{sugarFat}. CHK kinase-like protein is believed to respond to DNA damage by arresting the cell cycle. This same study also discovered an upregulation of genes related to peptide and carbohydrate processing in both high-fat and high-sugar diets, a finding the authors attributed to a general increase in caloric intake. Another recent Drosophila study investigated the transcriptomic effects of diets high in protein relative to sugar, diets high in sugar relative to protein, and diets with equal amounts of protein and sugar \cite{sugarProtein}. No differential expression was found in Drosophila melanogaster, a species known to thrive in all three dietary conditions. However, Drosophila mojavensisa and Drosophila arizonae showed substantial differential expression between the dietary conditions: Genes involved in carbohydrate and lipid metabolism were upregulated in response to high sugar low protein diets and genes involved in juvenile hormone (JH) and ecdysone were upregulated in reponse to low sugar high protein diets. Interestingly, prior studies have suggested that JH regulates body size by controlling ecdysone production, which modifies insulin signaling \cite{jHormone}. One transcriptomic study in honey bees found that pollen nutrition upregulates genes involved in macromolecule metabolism, longevity, and the insulin/TOR pathway required for physiological development \cite{alaux2}.

Likewise, studies on the transcriptomic effects of virus infection have been performed in insect models such as Drosophila and mosquito. These studies have shown that RNA silencing is a major antiviral strategy, but other pathways play substantial roles in many virus-host systems like the Toll pathway, IMD pathway, JAK/STAT pathway, Toll-7-autophagy pathway, and transcriptional pausing \cite{cherryReview, swevers}. It is important to note that general immune responses to viral infection in insects might be an indirect result of cellular damage \cite{swevers}. In fact, every virus-host interaction has its own particularities derived of the diverse ways of replication and infection cycle evolved by different viruses. In this sense, an intricate set of pro- and anti-virus host factors are involved in each case \cite{cherryReview, swevers}. For example, dicistroviruses Drosophila C virus (DCV) and cricket paralysis virus (CrPV) infection requires ribosomal proteins RpS6, RpL19 and RACK1, but such proteins are dispensable during infection by the rhabdovirus vesicular stomatitis virus (VSV) in Drosophila. In contrast, the autophagy pathway has proven to be an effective antiviral strategy against VSV, but has no role limiting DCV or other positive-sense RNA viruses \cite{swevers}. For an extensive up-to-date review of known antiviral and proviral factors for specific viruses in Drosophila see \cite{swevers}. Model insect studies have shed light on how honey bees respond to viruses because the function of many genes relies on the characterization of orthologue genes. Nonetheless, recent transcriptomic studies in honey bees have improved our understanding of compatible and incompatible virus-bee interactions. Key factors or hallmarks of the antiviral conserved defense pathways mentioned above have been found in some virus-bee systems (e.g. RNA silencing, Toll, IMD, JAK/STAT pathway, autophagy, and endocytosis) \cite{McMenamin}. Added to these, a non-sequence-specific antiviral response mediated by unspecific dsRNA pathway was discovered in honey bees \cite{Flenniken}. This path has been studied recently in more depth using transcriptomics, which found known immune factors but also new ones for which indispensability in antiviral response was determined by silencing \cite{Brutscher}. Similar to Drosophila examples, particular antiviral factors in honey bees are linked to specific viruses, specific developmental stages, the analyzed tissue, the route of inoculation, and the time (post-inoculation) during which the study was performed. This was clearly demonstrated when comparing honey bee responses to two related iflaviruses with very different infection dynamics, sacbrood bee virus (SBV) vs. deformed wing virus (DWV) \cite{Ryabov}. Authors observed differences in induction of defensin and hymenoptaecin immune-related genes, and suggested the results reflect adaptations to the different routes of transmission \cite{Ryabov}. In the case of dicistroviruses, few works have studied the impact of IAPV infection at transcriptional level. Chen et al. 2014 analyzed responses to IAPV infection in larvae and workers using microarrays \cite{Chen}. Many of the DEGs found were involved in immune response and energy-related metabolism, particularly in adults but not in brood. The authors propose this observed difference could be connected to latent infections in larvae (where host immunity is not perturbed) versus acute infections in adulthood (induced by stressors faced during development) \cite{Chen}. IAPV acute infection also alters the DNA methylation pattern of numerous genes that do not overlap the genes that are up- or down-regulated at the transcriptional level \cite{galbraith}. These works reiterate the conclusion that viruses trigger particular antiviral mechanisms by different means and depending on several factors. The honey bee antiviral pathways induced by specific viruses were recently reviewed \cite{McMenamin}; it is noteworthy to mention that many honey bee factors discovered by transcriptomics need further characterization to uncover their role in controlling (or promoting) viral infection in honey bees.

@@

There were several limitations in this study that could be improved upon in future studies. For instance, our comparison between the Galbraith data (single-drone colonies) and our data (polyandrous colonies) was limited by numerous extraneous variables between these studies. In addition to different molecular pipelines and bioinformatic preprocessing pipelines used between these studies, the Galbraith study focused on one-day old worker honey bees that were fed sugar and artificial pollen diet, whereas our study focused on adult worker honey bees that were fed bee-collected monofloral diets. Furthermore, the Galbraith data used eviscerated abdomens with attached fat bodies and only considered symptomatic honey bees for their infected treatment group, whereas we used whole bodies and considered both asymptomatic and symptomatic honey bees for our infected treatment group. There are also differences in the hours post inoculation and possible differences in the inoculation amount between the studies. Further differences between the studies can be found in their corresponding published methods sections \cite{adamInt, galbraith}. Our comparative visualization assessment between these two datasets was also somewhat limited because the virus effect in the Galbraith study used three replicates for each level, whereas the virus effect in our study used twelve replicates for each level that were actually further subdivided into six replicates for each diet level. Hence the apparent reduction in noise observed in the Galbraith data compared to our data in the PCA plots, parallel coordinate plots, scatterplot matrices, and litre plots may be an inadvertent product of the smaller number of replicates used and the lack of a secondary treatment group rather than solely the reduction in genetic variability through the single-drone colony design itself. With this in mind, while our current efforts may be a starting point, future studies can shed more light on signal:to:noise and differential expression differences between polyandrous colony designs and single-drone colony designs by controlling for extraneous factors more strictly than what we were able to do in the current line of work.
